Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-09-02 Manuscript/manuscript.Rmd
intro edits

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed the spatial organization of the human genome  [@lieberman2009comprehensive]. The genome has been shown to form tight, compact folds in order to fit inside the nucleus. This folding does not occur randomly; in fact, the genome exhibits a hierarchical 3D structure [@dixon2012topological]. Among this hierarchy are distict compartments where the genome folds and interacts with itself more frequently termed topologically associating domains (TADs). TADs have garnered much interest in recent years given that they are heavily linked to gene expression [@dixon2015chromatin]. Moreover, changing the 3D structure of the DNA, causing disruptions in the boundaries that demarcate TADs, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. It has been thought that the collection of particular functional genomic elements may contribute to the formation of TADs. Namely, the CCCTC-binding factor (CTCF) has been shown to be enriched at TAD boundaries [@ghirlando2016ctcf]. Additional transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries forming a cohesin ring under the proposed loop extrusion model, whereby loops are formed during interphase [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, chromatin has been shown to exhibit certain modifications at TAD boundaries, evidenced by the frequent observation of the H3K36me3 histone mark [@dixon2012topological]. The growing number of publically available genomic annotation data and their distinct patterns prompt the use of computational approaches toward TAD boundary prediction.

The classical approach is to build classification models that leverage the relationship between ChIP-seq genomic annotations and TAD boundary regions <!--[@mourad2016computational; @huang2015predicting; @hong2017computational]-->. A key factor in these approaches becomes resolution. Resolution determines the bin sizes of the linear genome used to enumerate contacts caused by chromatin folding. Due to high sequencing cost, most available Hi-C datasets have relatively low resolution, in the hundreds of kilobase range. This is in stark contrast to ChIP-seq experiments which annotate regions of the genome at hundreds of base pairs in resolution [??? REF, Bonev 2017?]. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation and its potential link to chromatin architecture. However, simply increasing sequencing depth to obtain higher resolution could have unintended consequences since this could introduce sparsity in the number of contacts which TAD callers are often ill-equiped to handle [@cresswell2019spectraltad, @sexton2012three]. Increased resolution could also further exacerbate the class imbalance created from the proportionally much smaller number of TAD boundary regions compared to non-TAD boundary regions throughout the linear genome. Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Additionally, identifying the most informative feature engineering steps used to quantify the relationships between genomic annotations and bins of varying resolutions remains an open question. Incorporating a distance measure has largely been ignored, but could offer a much more spatially descriptive feature space thereby improving predictive performance.

A method developed by Mourad et al [@mourad2016computational], called *HiCFeat*, used the percentage of overlap between several ChIP-seq defined transcription factor binding site (TFBS) regions and 1 kb genomic bins to build an L1-regularized multiple logistic regression model to predict TAD boundary regions. However, class imbalance was not addressed and instead, model performance was evaluated using area under the receiver operating characteristic curve (AUROC) which is known to be insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. <!--A more appropriate evaluation tool to consider would be a composite metric such as Matthews Correlation Coefficient (MCC), which penalizes for both false positives and false negatives.--> Furthermore, by only considering the percentage of overlap in defining the feature space, *HiCFeat* is limited in its granularity given that many regions on the linear genome will not be overlapped with any genomic annotations, resulting in a zero inflated model. This, likely being further exacerbated by the class imbalance. Two additional studies were proposed in 2015 and 2017 respectively, one using histone modifications in a Bayesian Additive Regression Trees (BART) model [@huang2015predicting] and the other using combined sets of TFBS, DNase I hypersensitive sites, and histone modifications together in an L2-based regularized linear model [@hong2017computational]. The BART model was built on relatively high resolution Hi-C data at 5 kb, while the L2-regularized model used 200 kb resolution Hi-C data. Both methods addressed class imbalance by performing random under-sampling (RUS), and used the elemental read count that appeared in each bin to describe the relationship between ChIP-seq annotations and genomic regions. Firstly, given the large disparity between the number of TAD vs non-TAD boundary regions, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Secondly, using read count overlap as the feature space suffers from similar limitations as the percentage of overlap. Moreover, it is unclear how either of these feature engineering steps compare to enumerating the distance in basepairs between ChIP-seq annotations and genomic bins in TAD boundary prediction.

@@

>>>>

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed important 3-dimensional (3D) spatial constructs within the human genome such as loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. Among these, TADs represent regions on the linear genome, at megabase scale, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics like gene regulation and cell differentiation, an example being the mechanism by which enhancers loop to come into contact with distal promoters to initiate transcription within these domains [@Lupianez:2015aa; @Franke:2016aa]. Likewise, disruption of TAD boundaries has been shown to lead to developmental disorders [@lupianez2016breaking], such as cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@Sun:2018aa; @Meaburn:2007aa]. Therefore, identifying the location of TAD boundaries remains a top priority in our goal to understand the functionality of the human genome.

A key factor in locating TAD boundaries is Hi-C data "resolution". Resolution determines the bin sizes of the linear genome used to enumerate contacts caused by chromatin folding. Lower resolutions contribute to larger bin sizes, which lead to variability in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. To increase the resolution of Hi-C data requires a quadratic increase in the total sequencing depth [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to define sub-domains and the boundaries that demarcate them. Furthermore, even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-Sequencing experiments used to annotate the genome, which ranges in the tens of base pairs. This resolution gap between 1D and 3D genomic maps can make it difficult to disentangle the role of additional factors related to TAD boundaries, thereby limiting our understanding of epigenomic regulation and its potential link to chromatin architecture. We hypothesize that, by leveraging genomic annotations--defined by high resolution ChIP-Seq data--known to be enriched at TAD boundaries, we may be able more precisely predict the locations of TAD boundaries on the linear genome.

Several methods have been developed to predict TAD boundary location using genomic annotations. However, most ignore key characteristics of 3D genomic data that are detrimental to model performance. A method developed by Mourad et al [@mourad2016computational], called *HiCFeat*, used the percentage of overlap between several ChIP-seq defined transcription factor binding site (TFBS) regions and 1 kb genomic bins to build an L1-regularized multiple logistic regression model to predict TAD boundary regions. However, the high level of resolution is likely to introduce heavily imbalanced classes created from the proportionally much smaller number of TAD boundary regions compared to non-TAD boundary regions throughout the linear genome. *HiCFeat* did not the address class imbalance. Instead, model performance was evaluated using area under the receiver operating characteristic curve (AUROC) which is known to be insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. <!--A more appropriate evaluation tool to consider would be a composite metric such as Matthews Correlation Coefficient (MCC), which penalizes for both false positives and false negatives.--> Furthermore, by only considering the percentage of overlap in defining the feature space, *HiCFeat* is limited in its granularity given that many regions on the linear genome will not be overlapped with any genomic annotations, resulting in a zero inflated model. <!--This, likely being further exacerbated by the class imbalance.--> Two additional studies were proposed in 2015 and 2017 respectively, one using histone modifications in a Bayesian Additive Regression Trees (BART) model [@huang2015predicting] and the other using combined sets of TFBS, DNase I hypersensitive sites, and histone modifications together in an L2-based regularized linear model [@hong2017computational]. The BART model was built on relatively high resolution Hi-C data at 5 kb, while the L2-regularized model used 200 kb resolution Hi-C data. Both methods addressed class imbalance by performing random under-sampling (RUS), and used the elemental read count that appeared in each bin to describe the relationship between ChIP-seq annotations and genomic regions. Firstly, given the large disparity between the number of TAD vs non-TAD boundary regions, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Secondly, using read count overlap as the feature space suffers from similar limitations as the percentage of overlap. Moreover, it is unclear how either of these feature engineering steps compare to enumerating the distance in basepairs between ChIP-seq annotations and genomic bins in TAD boundary prediction.

@@

<!--
Extensive research using high resolution maps (ranging from 750 bp to 1 mb) have probed much of the aspects of 3D chromatin architecture identifying several epigenomic modifications associated with TAD boundaries [@rao20143d; @bonev2017multiscale]. For instance, it is known that the CCCTC-binding factor (CTCF) is enriched at TAD boundaries acting as an insulator interfering with enhancer-promotor interactions [@ghirlando2016ctcf; @song2017ctcf; @ong2014ctcf]. Additional transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries forming a cohesin ring under the proposed loop extrusion model, whereby loops are formed during interphase [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. 
-->
